Large squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-22 large_number_squeezing_with_atoms.tex
Punctuation.
Tweak intro and add more references

Ultracold atomic interferometers are also reaching state of the art sensitivities in precision measurements \cite{robins:2013, altinET2013}. While all current atom interferometry is limited by technical noise sources, it appears that BEC-based coherent sources offer the best chance of overcoming some of these issues, for the same reasons that optical lasers are typically preferred for optical interferometry \cite{szigeti:2012}. If these technical noise sources can be controlled, atom interferometers will be limited by atomic shot noise, which leads to a sensitivity that scales as the square root of the atomic flux.   From that point, sensitivity will only be improved by increasing the atomic flux, and/or using non-classical quantum states such as squeezed states to go below the shot noise limit \cite{robins:2013}.  Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

@@

We wish to examine the scaling of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate.  Other system parameters such as the total number $N$ and the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular this fixes the values of $\theta$, $\phi$ and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but we will vary $\tau_\text{hold}$ to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The non-condensed fraction can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The non-condensed fraction $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the non-condensed fraction increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the non-condensed fraction \emph{decreases} with volume.  This suggests that the best system to use will be a three-dimensional box potential, and the system size can be increased until the condensate depletion becomes negligibly small.

@@

>>>>

Ultracold atomic interferometers are also reaching state of the art sensitivities in precision measurements \cite{robins:2013, altinET2013}. While all current atom interferometry is limited by technical noise sources, it appears that BEC-based coherent sources offer the best chance of overcoming some of these issues, for the same reasons that optical lasers are typically preferred for optical interferometry. In particular, it appears that Bose-condensed sources allow for higher fidelity mirrors and beamsplitters \cite{szigeti:2012}, are less sensitive to the effects of distortions in the optical wavefronts of the beamsplitters and mirrors and the Coriolis effect \cite{Debs:2011, robins:2013}, and are robust to the adverse effects of strong outcoupling when feedback-cooled \cite{semiclassicalstabilization, Szigeti:2009, Szigeti:2010}.

If these technical noise sources can be reduced and/or controlled, atom interferometers will be limited by atomic shot noise, which leads to a sensitivity that scales as the square root of the atomic flux. From that point, sensitivity will only be improved by increasing the atomic flux and brightness \cite{semiclassicallimit2007, outcouplercomparison, jeppesenET2008}, and/or using non-classical quantum states such as squeezed states to go below the shot noise limit \cite{robins:2013}.

Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

@@

We wish to examine the scaling of the non-condensed fraction $N_{\mathbf{k}\neq0}/N$ with the physical size of the condensate.  Other system parameters such as the total number $N$ and the occupation of the $\ket{a}$ and $\ket{b}$ states at $t_1$ ($n_a$ and $n_b$, respectively) are kept constant and chosen to maximise the squeezing of the equivalent two-mode model presented in Section~\ref{secTwoModeAnalytic}.  In particular, this fixes the values of $\theta$, $\phi$, and $\lambda_{ij} = \chi_{ij}\tau_\text{hold}$.  As the size of the condensate is scaled, $\hbar \chi_{ij} = U^\text{1D}_{ij}/L = U^\text{2D}_{ij}/A = U^\text{3D}_{ij}/V$ will change, but we will vary $\tau_\text{hold}$ to keep $\lambda_{ij}$ constant.  Therefore $\Lambda_a = n_a\chi_{aa}\tau_\text{hold}$ will also be constant.  With this in mind, we can write $\tau_\text{hold}$ in terms of the system size and the nonlinearity $U_{aa}$.  The non-condensed fraction can then be shown to scale with the physical size of the condensate as:
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \sqrt{L} \cdot \frac{1}{N} \sqrt{\frac{M U^\text{1D}_{aa} n_a}{2 \hbar^2 \pi^2}} \Lambda_a^{3/2} \cos^2\theta f^\text{1D}(\Lambda_a) &&\propto \sqrt{L}, \\
  \frac{N^\text{2D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx 1 \cdot \frac{1}{N} \frac{M U^\text{2D}_{aa} n_a }{4 \hbar^2} \Lambda_a \cos^2\theta f^\text{2D}(\Lambda_a) &&\propto \text{constant}, \\
  \frac{N^\text{3D}_{\mathbf{k}\neq 0}(t_3)}{N} &\approx \frac{1}{\sqrt{V}} \cdot \frac{1}{N} \left(\frac{M U^\text{3D}_{aa} n_a}{\hbar^2}\right)^{3/2} \frac{\Lambda_a^{1/2} \cos^2\theta f^\text{3D}(\Lambda_a)}{\sqrt{2} \pi^2} &&\propto \frac{1}{\sqrt{V}}.
\end{align}
The non-condensed fraction $N_{\mathbf{k}\neq 0}/N$ thus scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the non-condensed fraction increases with system size, whereas for 2D traps it is constant.  For three-dimensional traps, the non-condensed fraction \emph{decreases} with volume.  This suggests that the best system to use will be a three-dimensional box potential, and the system size can be increased until the condensate depletion becomes negligibly small.

@@

\bibitem{Debs:2011} Debs J E, Altin P A, Barter T H, D{\"{o}}ring D, Dennis G R, McDonald G, Anderson R P, Close J D, and
Robins N P 2011 Cold-atom gravimetry with a Bose-Einstein condensate {\emph Phys. Rev.} A {\bf 84} 033610

\bibitem{semiclassicalstabilization} Johnsson M T, Haine S A and Hope J J 2005 Stabilizing an atom laser using spatially selective pumping and feedback \emph{Phys. Rev.} A {\bf 72} 053603

\bibitem{Szigeti:2009} Szigeti S S, Hush M R, Carvalho A R R, and Hope J J 2009 Continuous measurement feedback control of a Bose-Einstein condensate using phase-contrast imaging \emph{Phys. Rev.} A {\bf 80} 013614

\bibitem{Szigeti:2010} S. S. Szigeti, M. R. Hush, A. R. R. Carvalho, and J. J. Hope 2010 Feedback control of an interacting Bose-Einstein condensate using phase-contrast imaging \emph{Phys. Rev.} A {\bf 82} 043632

\bibitem{semiclassicallimit2007} Johnsson M T, Haine S A, Hope J J, Robins N P, Figl C, Jeppesen M, Dugu{\'{e}} J and Close, J 2007 Semiclassical limits to the linewidth of an atom laser \emph{Phys. Rev.} A {\bf 75} 043618

\bibitem{outcouplercomparison} Dugu{\'{e}} J, Robins N P, Figl C, Jeppesen M, Johnsson M T, Hope J J, and Close J D 2007 Investigation and comparison of multi-state and two-state atom laser output-couplers \emph{Phys. Rev.} A {\bf 75} 053602

\bibitem{jeppesenET2008} Jeppesen M, Dugu{\'{e}}, Dennis G R, Johnsson M T, Figl C, Robins N P and Close J D 2008 Approaching the Heisenberg limit in an atom laser \emph{Phys. Rev.} A {\bf 77} 063618
